---
title: "Statin use and brain volume alterations"
author: "Mélissa Gentreau"
date: "2024-06-24"
output:
  pdf_document:
    latex_engine: xelatex
---
This code requires the packages *tidyverse*, *knitr*, *png*, *ggplot2*, *RColorBrewer*, *cowplot*, and *mediation*.       
```{r setup, include=FALSE}

library(tidyverse)
library(knitr)
library(png)
library(ggplot2)
library(RColorBrewer)
library(cowplot)
library(mediation)

load("~/Postdoc Uppsala/Analyses/Statin and Brain Aging/Longitudinal analyses/Imaging/envir_ukbMRI.RData")

```

### Function used to create the result table:     
```{r multivarlmR2, echo=TRUE}
multivarlmR2 <- function (data, outcome, varexpo, varajust, 
                          labelexpo = varexpo, decimal = 2) {
  # data        dataset
  # outcome     character vector of outcome names
  # varexpo     character vector of exposure variable (1)
  # varajust    character vector of covariates
  # labelexpo   character vector of labels of the exposure variable, colnames by default
  # decimal     number of digits after the decimal point, 2 by default
  
  x <- data[,c(varexpo)]
  res <- NULL
  
  for (i in 1:length(outcome)) {
    # y : variable i to explain
    y <- outcome[i]
    
    ## Empty matrix which will be used as a table
    mat_res <- matrix(NA, nrow = 0, ncol = 9)
    colnames(mat_res) <- c("Outcome","Exposure","Beta","SE",
                           "Lower CI", "Upper CI" ,"P value", "P.raw",
                           "Adjusted R-squared")
    
    varexpo <- c(varexpo)
    
    data2 <- na.omit(data[, c(y, varexpo, varajust)])
    
    
    # Writing the equation y ~ x1 + x2 +...+ xp
    formule <- paste(y, "~", varexpo)
    for (k in 1:length(varajust)) {formule <- paste(formule, "+", varajust[k])}
    formule <- as.formula(formule)
    
    # Model
    mod <- lm(formula = formule, data = data2)

    # Display results of the exposure variable only
    # i.e. the second row of coefficients and ICs (the first being the intercept)

    # Coefficients and standard error
    beta <- round(mod$coefficients, decimal)
    se <- round(summary(mod)$coefficients[, 2], decimal)
    CI <- round(confint(mod), decimal)
    
    # P-value extraction
    pval <- summary(mod)$coefficients[, 4]
    pvalr <- round(pval, 4)
    pvalr[as.double(pval) < 0.0001] <- "<0.0001"
    
    # Adjusted R-squared value
    adj.r.squared <- round(summary(mod)$adj.r.squared, 4)
    
    # If the exposure variable is a factor
    if (is.factor(x)) {
      
      lev <- levels(x)
      nlev <- nlevels(x)
      
      if (nlev == 2){
        line <- c(y, 
                   paste(varexpo, "_", lev[2], sep = ""),
                   beta[2], se[2], 
                   CI[2,1], CI[2,2], pvalr[2], pval[2], adj.r.squared)
        mat_res <- rbind(mat_res, line)
        
      } else if (nlev > 2) {
        line0 <- c(y, paste(varexpo, "_", lev[1], sep = ""), rep("", 6), adj.r.squared)
        mat_res <- rbind(mat_res, line0)
        
        for (k in 2:nlev) {
          line <- c(y, 
                     paste(varexpo, "_", lev[k], sep = ""),
                     beta[k], se[k], 
                     CI[k,1], CI[k,2], pvalr[k], pval[k], "")
          mat_res <- rbind(mat_res, line)
        }
      }
      
    } else if (is.numeric(x)){
      
      line <- c(y, varexpo, beta[2], se[2],
                 CI[2,1], CI[2,2], pvalr[2], pval[2], adj.r.squared)
      mat_res <- rbind(mat_res, line)
    }
    
    res <- rbind(res, mat_res)
    row.names(res) <- NULL
    
  }
  
  return(res)
  
}


```
## MAIN ANALYSES
```{r main, echo=TRUE}
imagsel <- c("GM_vol2", "WM_vol2", "cortical_vol2", "logWMH_vol2")

# Model 1
mmd1 <- multivarlmR2(data = as.data.frame(ukb2),
                    outcome = c(imagsel), 
                    varexpo = "statin0b",
                    varajust = c("centre2","age0","sex","ethnic0b","qualif0b", 
                                 "TDI0","APOE4","antidep2","ICV2"),
                    decimal = 2)

mmd1 <- as.data.frame(mmd1)
mmd1$Model <- rep("I",4)

# Model 2
mmd2 <- multivarlmR2(data = as.data.frame(ukb2),
                    outcome = c(imagsel), 
                    varexpo = "statin0b",
                    varajust = c("centre2","age0","sex","ethnic0b","qualif0b", 
                                 "TDI0","APOE4","antidep2","ICV2",
                                 "frq_alcohol0","smoking0","physact0b"),
                    decimal = 2)

mmd2 <- as.data.frame(mmd2)
mmd2$Model <- rep("II",4)

# Model 3
mmd3 <- multivarlmR2(data = as.data.frame(ukb2),
                    outcome = c(imagsel), 
                    varexpo = "statin0b",
                    varajust = c("centre2","age0","sex","ethnic0b","qualif0b", 
                                 "TDI0","APOE4","antidep2","ICV2",
                                 "frq_alcohol0","smoking0","physact0b",
                                 "BMI0","SBP0","DBP0","diabetes0","CHD0","stroke0",
                                 "headinjury0","depression0","insomn0"),
                    decimal = 2)

mmd3 <- as.data.frame(mmd3)
mmd3$Model <- rep("III",4)

res_mmd <- rbind(mmd1, mmd2, mmd3)
res_mmd <- res_mmd %>% arrange(Outcome)
res_mmd <- res_mmd[,-8]

res_mmd

```
       
### Model III according to statin type     

```{r main statintype, echo=TRUE}
ukb2$statintype0 <- as.character(ukb2$statin0b)
ukb2$statintype0[ukb2$fluvastatin0 == "1"] <- "Fluvastatin"
ukb2$statintype0[ukb2$pravastatin0 == "1"] <- "Pravastatin"
ukb2$statintype0[ukb2$rosuvastatin0 == "1"] <- "Rosuvastatin"
ukb2$statintype0[ukb2$atorvastatin0 == "1"] <- "Atorvastatin"
ukb2$statintype0[ukb2$simvastatin0 == "1"] <- "Simvastatin"

ukb2$statintype0 <- factor(ukb2$statintype0, 
                           levels = c("0","Simvastatin","Atorvastatin", 
                                      "Rosuvastatin","Pravastatin","Fluvastatin"))
table(ukb2$statintype0)


m.type <- multivarlmR2(data = as.data.frame(ukb2),
                       outcome = c(imagsel), 
                       varexpo = "statintype0",
                       varajust = c("centre2","age0","sex","ethnic0b","qualif0b", 
                                    "TDI0","APOE4","antidep2","ICV2",
                                    "frq_alcohol0","smoking0","physact0b",
                                    "BMI0","SBP0","DBP0","diabetes0","CHD0","stroke0",
                                    "headinjury0","depression0","insomn0"),
                       decimal = 2)

m.type <- as.data.frame(m.type)
m.type$Exposure <- rep(c("No statin","Simvastatin","Atorvastatin",
                         "Rosuvastatin","Pravastatin","Fluvastatin"), 4)
m.type <- m.type[,-8]
m.type

```
      
#### Figure: Associations between statin type and the volumes of grey matter, white matter, peripheral cortical grey matter and white matter hyperintensity. 
```{r main statintype plot, echo=TRUE, results='hide'}
m.type.p <- subset(m.type, m.type$Exposure != "No statin")
colnames(m.type.p)[5:7] <- c("Lower_CI","Upper_CI","P")
m.type.p$Beta <- as.numeric(m.type.p$Beta)
m.type.p$Lower_CI <- as.numeric(m.type.p$Lower_CI)
m.type.p$Upper_CI <- as.numeric(m.type.p$Upper_CI)
m.type.p$Exposure <- factor(m.type.p$Exposure, 
                            levels = c("Fluvastatin","Pravastatin","Rosuvastatin",
                                       "Atorvastatin","Simvastatin"))


p1 <- m.type.p %>% subset(m.type.p$Outcome == "GM_vol2") %>% 
  ggplot(aes(x = Exposure , y = Beta, color = Exposure)) + 
  geom_point(size = 0.9) +
  geom_hline(yintercept = 0, lty = 1, lwd = 1, color = "grey90") +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width=.2) +
  scale_color_discrete(type = rev(brewer.pal(n = 5, name = "Dark2"))) +
  xlab("") +
  ylab(expression(Beta)) +
  ylim(c(-20000,20000)) +
  ggtitle("Grey matter") +
  coord_flip() +
  guides(fill = "none", color = "none", linetype = "none", shape = "none") + 
  theme_minimal()
  

p2 <- m.type.p %>% subset(m.type.p$Outcome== "WM_vol2") %>% 
  ggplot(aes(x = Exposure, y = Beta, color = Exposure)) + 
  geom_point(size = 0.9) +
  geom_hline(yintercept = 0, lty = 1, lwd = 1, color = "grey90") +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width= 0.2) +
  scale_color_discrete(type = rev(brewer.pal(n = 5, name = "Dark2"))) +
  xlab("") +
  ylab(expression(Beta)) +
  ylim(c(-20000,20000)) +
  ggtitle("White matter") +
  coord_flip() +
  guides(fill = "none", color = "none", linetype = "none", shape = "none") + 
  theme_minimal()

p3 <- m.type.p %>% subset(m.type.p$Outcome == "cortical_vol2") %>% 
  ggplot(aes(x = Exposure, y = Beta, color = Exposure)) + 
  geom_point(size = 0.9)+
  geom_hline(yintercept = 0, lty = 1, lwd = 1, color = "grey90") +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width=.2) +
  scale_color_discrete(type = rev(brewer.pal(n = 5, name = "Dark2"))) +
  xlab("") +
  ylab(expression(Beta)) +
  ylim(c(-20000,20000)) +
  ggtitle("Peripheral cortical grey matter") +
  coord_flip() +
  guides(fill = "none", color = "none", linetype = "none", shape = "none") + 
  theme_minimal()


p4 <- m.type.p %>% subset(m.type.p$Outcome == "logWMH_vol2") %>% 
  ggplot(aes(x = Exposure, y = Beta, color = Exposure)) + 
  geom_point(size = 0.9)+
  geom_hline(yintercept = 0, lty = 1, lwd = 1, color = "grey90") +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = .2) +
  scale_color_discrete(type = rev(brewer.pal(n = 5, name = "Dark2"))) +
  xlab("") +
  ylab(expression(Beta)) +
  ylim(c(-1,1)) +
  ggtitle("White matter hyperintensities") +
  coord_flip() +
  guides(fill = "none", color = "none", linetype = "none", shape = "none") + 
  theme_minimal()

# plot_grid(p1, p2, p3, p4, ncol = 1)


```
     
## MEDIATION ANALYSIS     
### Statin, total cholesterol, and grey matter
```{r med GM, echo=TRUE}
GM <- lm(GM_vol2 ~ statin0b + chol0 + centre2 + age0 + sex + ethnic0b + qualif0b + 
           TDI0 + APOE4 + antidep2 + ICV2 + frq_alcohol0 + smoking0 + physact0b + 
           BMI0 + SBP0 + DBP0 + diabetes0 + CHD0 + stroke0 + headinjury0 + 
           depression0 + insomn0, 
         data = ukb2)

GM.med1 <- lm(chol0 ~ statin0b + centre2 + age0 + sex + ethnic0b + qualif0b + 
           TDI0 + APOE4 + antidep2 + ICV2 + frq_alcohol0 + smoking0 + physact0b + 
             BMI0 + SBP0 + DBP0 + diabetes0 + CHD0 + stroke0 + headinjury0 + 
             depression0 + insomn0,
              data = ukb2)


GM.chol <- mediate(GM.med1, GM, treat = "statin0b", mediator = "chol0", 
                      robustSE = T, sims = 1000)

summary(GM.chol)
```

### Statin, total cholesterol, and white matter
```{r med WM, echo=TRUE}
WM <- lm(WM_vol2 ~ statin0b + chol0 + centre2 + age0 + sex + ethnic0b + qualif0b + 
           TDI0 + APOE4 + antidep2 + ICV2 + frq_alcohol0 + smoking0 + physact0b + 
           BMI0 + SBP0 + DBP0 + diabetes0 + CHD0 + stroke0 + headinjury0 + 
           depression0 + insomn0, 
         data = ukb2)

WM.med1 <- lm(chol0 ~ statin0b + centre2 + age0 + sex + ethnic0b + qualif0b +
                TDI0 + APOE4 + antidep2 + ICV2 + frq_alcohol0 + smoking0 + physact0b +
                BMI0 + SBP0 + DBP0 + diabetes0 + CHD0 + stroke0 + headinjury0 + 
                depression0 + insomn0,
              data = ukb2)


WM.chol <- mediate(WM.med1, WM, treat = "statin0b", mediator = "chol0", 
                   robustSE = T, sims = 1000)
summary(WM.chol)
```

### Statin, total cholesterol, and peripheral cortical grey matter
```{r med pcGM, echo=TRUE}
cortical <- lm(cortical_vol2 ~ statin0b + chol0 + centre2 + age0 + sex + ethnic0b + 
                 qualif0b + TDI0 + APOE4 + antidep2 + ICV2 + frq_alcohol0 + smoking0 + 
                 physact0b + BMI0 + SBP0 + DBP0 + diabetes0 + CHD0 + stroke0 + 
                 headinjury0 + depression0 + insomn0, 
          data = ukb2)

cortical.med1 <- lm(chol0 ~ statin0b + centre2 + age0 + sex + ethnic0b + 
                 qualif0b + TDI0 + APOE4 + antidep2 + ICV2 + frq_alcohol0 + smoking0 + 
                 physact0b + BMI0 + SBP0 + DBP0 + diabetes0 + CHD0 + stroke0 + 
                 headinjury0 + depression0 + insomn0,
               data = subset(ukb2, is.na(ukb2$cortical_vol2) == F))


cortical.chol <- mediate(cortical.med1, cortical, treat = "statin0b", mediator = "chol0", 
                    robustSE = T, sims = 1000)
summary(cortical.chol)
```

### Statin, total cholesterol, and WMH
```{r med WMH, echo=TRUE}
WMH <- lm(logWMH_vol2 ~ statin0b + chol0 + centre2 + age0 + sex + ethnic0b + qualif0b + 
            TDI0 + APOE4 + antidep2 + ICV2 + frq_alcohol0 + smoking0 + physact0b + 
            BMI0 + SBP0 + DBP0 + diabetes0 + CHD0 + stroke0 + headinjury0 + 
            depression0 + insomn0, 
          data = ukb2)

WMH.med1 <- lm(chol0 ~ statin0b + centre2 + age0 + sex + ethnic0b + qualif0b +
                 TDI0 + APOE4 + antidep2 + ICV2 + frq_alcohol0 + smoking0 + physact0b + 
                 BMI0 + SBP0 + DBP0 + diabetes0 + CHD0 + stroke0 + headinjury0 + 
                 depression0 + insomn0,
                data = subset(ukb2, is.na(ukb2$logWMH_vol2) == F))


WMH.chol <- mediate(WMH.med1, WMH, treat = "statin0b", mediator = "chol0", 
                          robustSE = T, sims = 1000)
summary(WMH.chol)

```
      
## SECONDARY ANALYSIS      
```{r subreg, echo=TRUE}
# Cortical structures with FAST and sub-cortical structures with FIRST
imag <- colnames(ukb2)[c(337:360,362:371,373:380,382:388,390:392,394:396)]
# length(imag)
# 55 imaging variables


# Model 3
mod3 <- multivarlmR2(data = as.data.frame(ukb2),
                    outcome = imag, 
                    varexpo = "statin0b",
                    varajust = c("centre2","age0","sex","ethnic0b","qualif0b", 
                                    "TDI0","APOE4","antidep2","ICV2",
                                    "frq_alcohol0","smoking0","physact0b",
                                    "BMI0","SBP0","DBP0","diabetes0","CHD0","stroke0",
                                    "headinjury0","depression0","insomn0"),
                    decimal = 2)


mod3 <- as.data.frame(mod3)
mod3$P.adjust <- p.adjust(mod3$P.raw, method = "fdr")
mod3 <- mod3[,c("Outcome","Beta","Lower CI","Upper CI",
                "P value","P.adjust","Adjusted R-squared")]

mod3

```
      
## SENSITIVITY ANALYSIS
### Never users versus long-term statin users        
```{r statinlongterm, echo=TRUE}
## Create a variable for long-term statin users

ukb2$statin_t02 <- rep(NA, dim(ukb2)[1])
# 0: never used statin
ukb2$statin_t02[ukb2$statin0b == 0 & ukb2$statin2b == 0] <- 0 
# 1: started using statin at the imaging visit (removed)  
ukb2$statin_t02[ukb2$statin0b == 0 & ukb2$statin2b == 1] <- 1
# 2: reported using statin at recruitment but not at the imaging visit (removed)
ukb2$statin_t02[ukb2$statin0b == 1 & ukb2$statin2b == 0] <- 2
# 3: continuously used statin between recruitment and imaging visit
ukb2$statin_t02[ukb2$statin0b == 1 & ukb2$statin2b == 1] <- 3

ukb2$statin_t02 <- as.factor(ukb2$statin_t02)
table(ukb2$statin_t02)

ukb3 <- ukb2 %>% filter(ukb2$statin_t02 == "0" | ukb2$statin_t02 == "3")
table(ukb3$statin0b)

```
```{r sensitivity  longt, echo=TRUE}

# Model 1
m.longt1 <- multivarlmR2(data = as.data.frame(ukb3),
                         outcome = imagsel, 
                         varexpo = "statin0b",
                         varajust = c("centre2","age0","sex","ethnic0b","qualif0b", 
                                      "TDI0","APOE4","antidep2","ICV2"),
                         decimal = 2)

m.longt1 <- as.data.frame(m.longt1)
m.longt1$Model <- rep("I",4)

# Model 2
m.longt2 <- multivarlmR2(data = as.data.frame(ukb3),
                         outcome = imagsel, 
                         varexpo = "statin0b",
                         varajust = c("centre2","age0","sex","ethnic0b","qualif0b", 
                                      "TDI0","APOE4","antidep2","ICV2",
                                      "frq_alcohol0","smoking0","physact0b"),
                         decimal = 2)

m.longt2 <- as.data.frame(m.longt2)
m.longt2$Model <- rep("II",4)

# Model 3
m.longt3 <- multivarlmR2(data = as.data.frame(ukb3),
                         outcome = imagsel, 
                         varexpo = "statin0b",
                         varajust = c("centre2","age0","sex","ethnic0b","qualif0b", 
                                      "TDI0","APOE4","antidep2","ICV2",
                                      "frq_alcohol0","smoking0","physact0b","BMI0",
                                      "SBP0","DBP0","diabetes0","CHD0","stroke0",
                                      "headinjury0","depression0","insomn0"),
                         decimal = 2)

m.longt3 <- as.data.frame(m.longt3)
m.longt3$Model <- rep("III",4)

m.longt <- rbind(m.longt1, m.longt2, m.longt3)
m.longt <- m.longt %>% arrange(Outcome)
m.longt <- m.longt[,-8]

m.longt
```

### Replace self-reported medical history variables with variables based on ICD10 diagnoses
```{r sensitivity data, echo=FALSE, include=FALSE, eval=TRUE}

icdv <- read_csv("~/Postdoc Uppsala/UK Biobank/Generated variables/ukb30172_676881_ICD10_PREV.csv")

ukb2 <- ukb2 %>% inner_join(icdv, by = "eid_30172") 

ukb2  <- ukb2 %>% mutate_at(c(408:413), factor)
colnames(ukb2)[c(409,410,413)] <- c("CHD_ICD0", "stroke_ICD0", "sleepdis_ICD0")

```


```{r sensitivity  ICD10, echo=TRUE}

mod3ICD <- multivarlmR2(data = as.data.frame(ukb2),
                        outcome = imagsel, 
                        varexpo = "statin0b",
                        varajust = c("centre2","age0","sex","ethnic0b","qualif0b", 
                                     "TDI0","APOE4","antidep2","ICV2",
                                     "frq_alcohol0","smoking0","physact0b","BMI0",
                                     "SBP0","DBP0","diabetesICD0","CHD_ICD0",
                                     "stroke_ICD0","headinjuryICD0","depressionICD0"),
                    decimal = 2)

mod3ICD <- as.data.frame(mod3ICD)
mod3ICD$Model <- rep("III",4)
mod3ICD <- mod3ICD[,-8]

mod3ICD

```


